---
title: "Bacterial community complexity in the phyllosphere penalises specialists over generalists - Results Section 3"
author: "Rudolf Schlechter"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

library(here)
library(tidyverse)
library(rstatix)
library(ggh4x)
library(patchwork)

source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))

# Data
source(here('code', 'spat_K_fractions_auc.R'))
source(here('code', 'function_fractional_change_median.R'))
K_auc_aggregation <- K_auc_fold_change %>% filter(type == "aggregate_fraction")

set.seed(19900725)
```

## Effect of community complexity on intraspecific spatial relations

```{r}
K_fractions %>% head
```

Community context was expected to influence the spatial distribution patterns (aggregation, randomness, regularity) within bacterial populations in the phyllosphere. To evaluate this, we first determined relative frequencies of a spatial pattern based on $\hat{K}$(*r*), for each bacterial strain within every community context. Subsequently, we quantified the area under the curve of each spatial pattern and calculated the fractional change compared to the near-isogenic control condition, C.


```{r analysis_type}
## Summary
summary_type <- fractional_change_median(K_auc_fold_change, group = "type")

## Kruskal-Wallis
kw_type <- K_auc_fold_change %>% 
    kruskal_test(fractional_change ~ type) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

kw_eff_type <- K_auc_fold_change %>% 
    kruskal_effsize(fractional_change ~ type, ci = TRUE, nboot = 100)

## One-sample Wilcoxon test
w1_type <-  K_auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

```

Our initial analysis showed that spatial distribution patterns within populations differed from their respective controls. Combined, there was a reduction of `r summary_type$percentage[summary_type$type=="aggregate_fraction"]`% and `r summary_type$percentage[summary_type$type=="regular_fraction"]`% in aggregation and regularity, respectively, while random distributions increased by `r summary_type$percentage[summary_type$type=="random_fraction"]`%. Increase in randomness was observed mainly in *Methylobacterium*.

```{r analysis_dpi}
## Summary
summary_dpi <- fractional_change_median(K_auc_fold_change, group="dpi")

## One sample Wilcoxon test
w1_dpi = K_auc_fold_change %>% 
    group_by(dpi, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(dpi, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

## Two samples Wilcoxon test
w2_dpi <- K_auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ dpi, p.adjust.method = "holm", detailed = TRUE)
```

The observed distribution patterns were consistent across the two time points. That is, aggregation and regularity within populations decreased, while randomness increased in 7 and 14 dpi. Given that aggregation and regularity followed a consistent pattern, and randomness is a factor that would be related to neutral rather than deterministic processes, we decided to focus our analysis only on changes in aggregation within populations and the factors influencing this spatial pattern.

```{r analysis_taxa_strain}
## Summary
summary_taxa <- fractional_change_median(K_auc_aggregation, 
                                         group=c("taxa", "dpi"))
summary_strain <-  fractional_change_median(K_auc_aggregation, 
                                            group = c("strain", "dpi")) %>% 
    filter(dpi == "07dpi")

## Two samples Wilcoxon test for taxa (Methylobacterium vs Sphingomonas)
w2_taxa <- K_auc_aggregation %>% 
    group_by(dpi, synID) %>% 
    wilcox_test(fractional_change ~ taxa, p.adjust.method = "holm", detailed = TRUE)

## One sample Wilcoxon test for taxa (Methylobacterium vs Sphingomonas)
w1_taxa <-  K_auc_aggregation %>% 
    group_by(synID, dpi, taxa) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(synID, dpi, taxa, estimate, statistic, p) %>% 
    mutate(
        symbol = ifelse(p < 0.05, "*", ""),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

w1_taxa_plot_data <- fractional_change_median(K_auc_aggregation, group = c("synID", "dpi", "taxa")) %>% 
    left_join(., w1_taxa, by = c("synID", "dpi", "taxa"))

## One sample Wilcoxon test for strains
w1_strain  <-  K_auc_aggregation %>% 
    group_by(synID, dpi, strain) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(synID, dpi, strain, estimate, statistic, p) %>% 
    mutate(
        symbol = ifelse(p < 0.05, "*", ""),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

w1_strain_plot_data <- fractional_change_median(K_auc_aggregation, group = c("synID", "dpi", "strain")) %>% 
    left_join(., w1_strain, by = c("synID", "dpi", "strain"))

```


```{r analysis_synID_strain}
##  Summary
summary_synID <- K_auc_aggregation %>% 
    group_by(dpi, synID) %>% 
    summarise(median = median(fractional_change),
              q1 = format(round(quantile(fractional_change, 0.25), 2), nsmall = 2),
              q3 = format(round(quantile(fractional_change, 0.75), 2), nsmall = 2),
              IQR = paste0(q1,"-(",q3,")", sep=''),
              percentage = abs(100*median),
              .groups = "drop")

summary_synID_strain <- K_auc_aggregation %>% 
    filter(dpi == "07dpi") %>% 
    group_by(synID, strain) %>% 
    summarise(median = median(fractional_change),
              q1 = format(round(quantile(fractional_change, 0.25), 2), nsmall = 2),
              q3 = format(round(quantile(fractional_change, 0.75), 2), nsmall = 2),
              IQR = paste0(q1,"-(",q3,")", sep=''),
              percentage = abs(100*median),
              .groups = "drop")

## Two samples Wilcoxon test for SynCom ID (S2 vs S3)
w2_synID <- K_auc_aggregation %>% 
    group_by(dpi) %>% 
    wilcox_test(fractional_change ~ synID, p.adjust.method = "holm")

##  Dunn test for aggregation patterns of each strain at 7 dpi
d_synID_strain <- K_auc_aggregation %>% 
    filter(dpi == "07dpi") %>% 
    group_by(synID) %>% 
    dunn_test(fractional_change ~ strain, p.adjust.method = "holm") %>% 
    filter(p.adj < 0.05)

```

When comparing the aggregation between bacterial taxa, we identified differences between *Methylobacterium* and *Sphingomonas* only at 7 dpi (Wilcoxon, *W* = `r w2_taxa$statistic[w2_taxa$dpi=="07dpi"]`, *p* = `r w2_taxa$p[w2_taxa$dpi=="07dpi"]`). At this time point, aggregation differed from S2 to S3 (Wilcoxon, *W* = `r w2_synID$statistic[w2_synID$dpi=="07dpi"]`, *p* = `r w2_synID$p[w2_synID$dpi=="07dpi"]`). While both taxa exhibited decreased self-aggregation, *Methylobacterium* showed the largest decrease in aggregation in S3 (`r w1_taxa_plot_data$percentage[w1_taxa_plot_data$synID=="S3"&w1_taxa_plot_data$dpi=="07dpi"&w1_taxa_plot_data$taxa=="Methylobacterium"]`%) relative to the control. This decrease was reflected in a reduction of `r summary_synID_strain$percentage[summary_synID_strain$synID=="S3"&summary_synID_strain$strain=="meL85"]`%, `r summary_synID_strain$percentage[summary_synID_strain$synID=="S3"&summary_synID_strain$strain=="meL92"]`%, and `r summary_synID_strain$percentage[summary_synID_strain$synID=="S3"&summary_synID_strain$strain=="mr01"]`% in aggregation for populations of MeL85, MeL92, and Mr0-1, respectively. Conversely, the sphingomonads SmFR1 and SpFA2 were less affected, with reductions in self-aggregation of `r summary_synID_strain$percentage[summary_synID_strain$synID=="S3"&summary_synID_strain$strain=="smfr1"]`% and `r summary_synID_strain$percentage[summary_synID_strain$synID=="S3"&summary_synID_strain$strain=="spfa2"]`% in S3, respectively. Consequently, MeL85 and MeL92 were found to be statistically different to SmFR1 within S3.

The results of our analysis indicate that bacterial taxa respond differentially to community complexity, resulting in decreased self-aggregation and self-regularity, alongside an increase in random distribution patterns. This was particularly the case for *Methylobacterium* at 7 dpi in three-species communities. Furthermore and consistent with our results at the CFU and observed cell densities, aggregation was most affected in methylobacteria compared to sphingomonads, particularly MeL85 and MeL92 were negatively affected, while SmFR1 was least affected.



## Plots

```{r plot, include = FALSE}
plt.frq <- K_fractions %>% 
    group_by(taxa, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(x=r, y=mean_fraction))+
    facet_nested(fct_relevel(taxa, c("Sphingomonas", "Methylobacterium")) ~ dpi + synID, 
                 nest_line = element_line(color="black"),
                 labeller = labeller(dpi = dpi.lab2))+
    geom_area(aes(fill=fct_reorder(type, index)), color = "black", linewidth = 0.25)+
    scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                       expand = c(0,0), 
                       limit=c(0.2,30), 
                       breaks = seq(5, 25, 10))+
    scale_y_continuous(name = bquote("Relative frequency of"~hat(K)(italic(r))), 
                       expand = c(0,0), 
                       breaks = seq(0,1,0.5))+
    scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
    theme_rs()+
    theme(aspect.ratio = 1, 
          panel.spacing.x = grid::unit(0.25, "line"),
          panel.spacing.y = grid::unit(0.5, "line"),
          strip.text.x = element_text(margin = margin(b = 1, t = 5), face = "plain"),
          strip.text.y = element_text(angle=0, vjust = 0.5, hjust = 0),
          legend.position = "bottom")

plt.taxa <- w1_taxa_plot_data %>% 
    ggplot(aes(synID, taxa))+
    facet_wrap(~dpi, ncol = 2, labeller = labeller(dpi = dpi.lab2))+
    geom_tile(aes(fill = median*100), colour= "black")+
    geom_text(aes(label = symbol), vjust = 0.77, size = 5)+
    scale_fill_gradientn(name = "Change in\naggregation (%)",
        colours = rev(brewer.pal(7, "Blues")), 
        values=c(0, 0.6, 1), 
        limits = c(-70, 0), 
        breaks = seq(-70, 0, 20))+
    labs(x="", y="")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust=0.5, vjust=3),
          axis.text.y = element_text(face="italic"),
          strip.text = element_text(face="plain"),
          legend.position = "bottom")

plt.strain <- w1_strain_plot_data %>% 
    ggplot(aes(synID, strain))+
    facet_wrap(~dpi, ncol = 2, labeller = labeller(dpi = dpi.lab2))+
    geom_tile(aes(fill = median*100), colour= "black")+
    geom_text(aes(label = symbol), vjust = 0.77, size = 5)+
    scale_fill_gradientn(name = "Change in\naggregation (%)",
        colours = rev(brewer.pal(7, "Blues")), 
        values=c(0, 0.6, 1), 
        limits = c(-70, 0), 
        breaks = seq(-70, 0, 20))+
    scale_y_discrete(name = "Focal strain", labels = sp.lab)+
    scale_x_discrete(name="SynCom")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust = 0.5, vjust = 3),
          strip.text = element_text(face="plain"),
          legend.position = "bottom")

```

```{r figure_5, fig.id=TRUE, fig.cap="Bacterial population density in the arabidopsis phyllosphere", fig.width=3.5, fig.height=2.5, out.width="50%"}

# Plot
p2 <- wrap_plots(plt.taxa, plt.strain, ncol = 1, guides = "collect") +
    plot_layout(heights = c(1,3)) &
    theme(legend.box.just = "left",
          legend.position = "bottom",
          legend.margin = margin(-5,0,5,0)) &
    guides(fill = guide_colorbar(title.position="top", title.hjust = 0.5, barheight = unit(0.1, 'in')))


wrap_elements(full = plt.frq) + p2 +
    plot_annotation(tag_levels = "A") + 
    plot_layout(widths = c(3,1)) &
    theme(panel.spacing.x = unit(0.5, "lines"),
          plot.margin = margin(1,2,0,2),
          plot.tag = element_text(size = 7))

# Save plot
#ggsave(here("results", "figure_5.png"), width = 6.89, height = 2.5)
#ggsave(here("results", "figure_5.tiff"), width = 6.89, height = 2.5)

```

# Supplemental Figure

```{r plot_sup_freq_all, include=FALSE}
fractions_strain <- K_fractions %>% 
    group_by(strain, syncom, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3))

freq_plots <- function(name){
    plt <- fractions_strain %>% 
        filter(strain == name) %>% 
        ggplot(aes(x=r, y=mean_fraction))+
        facet_nested(synID + syncom ~ dpi, 
                     nest_line = element_line(color="black"),
                     labeller = labeller(dpi=dpi.lab2))+
        geom_area(aes(fill=fct_reorder(type, index)), color = "black", linewidth = 0.25)+
        scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                           expand = c(0,0), 
                           limit=c(0.2,30), 
                           breaks = seq(10, 20, 10))+
        scale_y_continuous(name = bquote("Relative frequency of"~hat(K)(italic(r))), 
                           expand = c(0,0), 
                           breaks = seq(0,1,0.5))+
        scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
        theme_rs()+
        theme(panel.spacing.x = grid::unit(0.25, "line"),
              panel.spacing.y = grid::unit(0.5, "line"),
              strip.text = element_text(angle=0, vjust = 0.5, hjust = 0, 
                                          margin = margin(r = 5, l = 2), face = "plain"))
    
    return(plt)
}

frq.plt.meL85 <- freq_plots("meL85")
frq.plt.meL92 <- freq_plots("meL92")
frq.plt.mr01 <- freq_plots("mr01")
frq.plt.smfr1 <- freq_plots("smfr1")
frq.plt.spfa2 <- freq_plots("spfa2")

```

```{r figure_s3, warning=FALSE, fig.id=TRUE, fig.cap="Frequency plots K", fig.width=6.89, fig.height=5, out.width="75%"}

# Set up plot
frq.plt.meL85 + frq.plt.meL92 + frq.plt.mr01 + frq.plt.smfr1 + frq.plt.spfa2 + 
    plot_annotation(tag_levels = "A")+
    plot_layout(ncol = 5, guides = "collect") & 
    theme(legend.position = "bottom",
          axis.text = element_text(size = 5),
          axis.title = element_text(size = 6),
          strip.text = element_text(size = 6),
          plot.tag = element_text(size = 7),
          plot.margin = margin(1,0,1,0))

# Save plot
#ggsave(here("results", "figure_s3.png"), width = 6.89, height = 5)
#ggsave(here("results", "figure_s3.tiff"), width = 6.89, height = 5)

```

```{r figure_s4, fig.id=TRUE, fig.cap="Spatial patterns", fig.width=3.46, fig.height=3, out.width="50%"}
plt.pattern <- K_auc_fold_change %>% 
    ggplot(aes(taxa, fractional_change))+
    facet_grid(dpi ~ type, labeller = labeller(dpi = dpi.lab2, type = pattern.lab))+
    geom_jitter(aes(fill = strain), color = "black", size = 1, stroke = 0.25, width = 0.1, alpha = 0.8, pch = 21)+
    geom_boxplot(alpha = 0.5, fill = "white", width = 0.2, outlier.alpha = 0, linewidth = 0.25)+
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.25)+
    theme_rs()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face="italic"),
          strip.text = element_text(face = "plain"),
          panel.spacing.x = grid::unit(0.5, "line"),
              panel.spacing.y = grid::unit(0.5, "line"))+
    labs(x = "", y = "Fractional change in relation to C")+
    scale_fill_manual(name = "Strain", values=sp.pal, labels=sp.lab)

# Save plot
plt.pattern
#ggsave(here("results", "figure_s4.png"), width = 3.46, height = 3)
#ggsave(here("results", "figure_s4.tiff"), width = 3.46, height = 3)

```

