---
title: "Bacterial community complexity in the phyllosphere penalises specialists over generalists - Results Section 4"
author: "Rudolf Schlechter"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

library(here)
library(tidyverse)
library(rstatix)
library(patchwork)
library(ggh4x)

source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))

#Data
source(here('code', 'spat_pcf_fractions_auc.R'))
source(here('code', 'function_fractional_change_median.R'))
G_auc_fold_change <- G_auc_fold_change %>% 
    mutate(taxa_type = case_when(
        taxa_pair == "MM" | taxa_pair == "SS" ~ "within",
        taxa_pair == "MS" ~ "between"))
G_auc_aggregation <- G_auc_fold_change %>% filter(type == "aggregate_fraction")


set.seed(19900725)
```

### Effect of community complexity on interspecific spatial correlations

We investigated the effect of community complexity on the spatial patterns between pairs of species. Pair cross correlations, $\hat{g}$(*r*), indicated whether an aggregated, random, or regular pattern was detected and how frequent for a pair at a given distance. Subsequently, we determined the area under the curve to compare spatial patterns between species pairs in S3 relative to S2. We considered S2 as a control, since S2 represents the pair of species without any other competitor present.

```{r analysis_type}
## Summary
summary_type <- fractional_change_median(G_auc_fold_change, group = "type")

## Kruskal-Wallis
kw_type <- G_auc_fold_change %>% 
    kruskal_test(fractional_change ~ type) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

kw_eff_type <- G_auc_fold_change %>% 
    kruskal_effsize(fractional_change ~ type, ci=TRUE, nboot=100)

## Dunn's test
dunn_type <- G_auc_fold_change %>% 
    dunn_test(fractional_change ~ type, p.adjust.method = "holm")

## One-sample Wilcoxon test
w1_type <-  G_auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

```

In general, we observed that pairs of species aggregated within S2 communities at each sampling point (Figure 6A). However, the presence of a third strain decreased these aggregation frequencies, subsequently altering regular and random patterns within S3 (Kruskal-Wallis, *H*(`r kw_type$df`) = `r kw_type$statistic`, *p* `r kw_type$p_label`). Specifically, aggregation and regularity decreased by `r summary_type$percentage[summary_type$type=="aggregate_fraction"]`% and `r summary_type$percentage[summary_type$type=="regular_fraction"]`%, respectively, while randomness increased by `r summary_type$percentage[summary_type$type=="random_fraction"]`%.

```{r analysis_dpi}
## Summary
summary_dpi <- fractional_change_median(G_auc_fold_change, group=c("dpi", "type"))

## One sample Wilcoxon test
w1_dpi = G_auc_fold_change %>% 
    group_by(dpi, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(dpi, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

## Two samples Wilcoxon test
w2_dpi <- G_auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ dpi, p.adjust.method = "holm", detailed=TRUE)
```

To further dissect these differences, we grouped the species pairs based on their identity, as a pair was composed of either two *Methylobacterium* (MM), a *Methylobacterium* and a *Sphingomonas* (MS), or two *Sphingomonas* (SS) species. We noticed that randomness was notably high at 7 dpi in pairs within S3 community (One-sample Wilcoxon, *W* = `r w1_dpi$statistic[w1_dpi$type=="random_fraction" & w1_dpi$dpi=="07dpi"]`, *p* `r w1_dpi$p_label[w1_dpi$type=="random_fraction" & w1_dpi$dpi=="07dpi"]`), with this increase being particularly pronounced in pairs comprising at least one *Methylobacterium* species (MM and MS). However, the elevated randomness declined over time, becoming comparable to S2 at 14 dpi. The reduction in aggregation was primarily observed in MM pairs at 14 dpi. By contrast, pairs exclusively composed of *Sphingomonas* species (SS) were not different between S2 and S3.

Our findings collectively indicate that *Methylobacterium* species are more susceptible than *Sphingomonas* to changes in their spatial distribution depending on their community context. This susceptibility is reflected by pairs containing at least one *Methylobacterium* species displaying random distribution patterns in more complex communities, implying that *Methylobacterium* tend to form fewer stable associations than *Sphingomonas*.

```{r analysis_taxa_strain}
## One sample Wilcoxon test for taxa pairs
w1_taxa <-  G_auc_fold_change %>% 
    group_by(dpi, type, taxa_pair) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(dpi, type, taxa_pair, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "*", TRUE ~ ""))
w1_taxa %>% filter(p_label == "*")

## Kruskal-Wallis test for taxa pairs
kw_taxa <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ taxa_pair)

dunn_taxa <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    dunn_test(fractional_change ~ taxa_pair, p.adjust.method = "holm", detailed = TRUE) %>% 
    filter(p.adj < 0.05)

## Two samples Wilcoxon  test for taxa pairs over time
w2_dpi_taxa <- G_auc_fold_change %>% 
    group_by(type, taxa_pair) %>% 
    wilcox_test(fractional_change ~ dpi, p.adjust.method = "holm", detailed = TRUE)

```


```{r, include = FALSE}
## Kruskal Wallis test for strain pairs
kw_strain <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ strain_pair)

dunn_strain <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    dunn_test(fractional_change ~ strain_pair, p.adjust.method = "holm", detailed = TRUE) %>% 
    dplyr::select(dpi, type, group1, group2, p, p.adj, p.adj.signif)

## Two samples Wilcoxon test for strain pairs
w2_strain <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ strain_pair, p.adjust.method = "holm", detailed = TRUE)
w2_strain %>% filter(p.adj < 0.05)

## One sample Wilcoxon test for strain pairs
w1_strain  <-  G_auc_fold_change %>% 
    group_by(dpi, strain_pair, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, p.adjust.method = "holm", detailed = TRUE) %>% 
    dplyr::select(dpi, strain_pair, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))
w1_strain %>% filter(p < 0.05)
```

```{r, include = FALSE}
## Kruskal Wallis test for  Syncom
kw_syncom <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ syncom)

## Two samples Wilcoxon test for SynCom
w2_syncom <- G_auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ syncom, p.adjust.method = "holm", detailed = TRUE) %>% 
    filter(p.adj < 0.05)

## One sample Wilcoxon test for SynCom
w1_syncom <-  G_auc_fold_change %>% 
    group_by(dpi, type, syncom) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    dplyr::select(dpi, type, syncom, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "*", TRUE ~ as.character(p)))
w1_syncom %>% filter(p_label == "*")
```

## Plots

```{r, include = FALSE}
plt.frq <- G_fractions %>% 
    group_by(taxa_pair, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(x=r, y=mean_fraction))+
    facet_nested(taxa_pair ~ dpi + synID, 
                 nest_line = element_line(color="black"),
                 labeller = labeller(dpi = dpi.lab2))+
    geom_area(aes(fill=fct_reorder(type, index)), color = "black", linewidth = 0.25)+
    scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                       expand = c(0,0), limit=c(0.2,30), breaks = seq(5, 25, 10))+
    scale_y_continuous(name = bquote("Relative frequency of"~hat(g)(italic(r))),
                       expand = c(0,0), breaks = seq(0,1,0.5))+
    scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
    theme_rs()+
    theme(panel.spacing.x = grid::unit(0.25, "line"),
          panel.spacing.y = grid::unit(0.5, "line"),
          strip.text.x = element_text(margin = margin(b = 1, t = 5), face = "plain"),
          strip.text.y = element_text(angle=0, vjust = 0.5, hjust = 0, face = "plain"),
          legend.position = "bottom")

plt.pattern <- left_join(G_auc_fold_change, w1_taxa, by = c("dpi", "type", "taxa_pair")) %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(taxa_pair, fractional_change))+
    geom_text(aes(y=2.5, label=p_label), size = 3, parse=FALSE)+
    facet_grid(dpi ~ fct_reorder(type, index), labeller = labeller(.rows = dpi.lab2, .cols = pattern.lab))+
    geom_jitter(color = "black", size = 1, stroke = 0.25, width = 0.1, alpha = 0.8)+
    geom_boxplot(alpha = 0.5, fill = "white", width = 0.2, outlier.alpha = 0, linewidth = 0.25)+
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.25)+
    theme_rs()+
    theme(strip.text = element_text(face = "plain"),
          panel.spacing.x = grid::unit(0.25, "line"),
          panel.spacing.y = grid::unit(0.25, "line"))+
    labs(x = "Taxa pair", y = "Fractional change relative to S2")

```

```{r figure_6, fig.id=TRUE, dev = "png", fig.width=6.89, fig.height=2.5, out.width="75%"}
# Set up plot
plt.frq + wrap_elements(plot = plt.pattern) + 
    plot_annotation(tag_levels = "A") &
    theme(legend.box.just = "center",
          plot.margin = margin(0,1,0,1),
          plot.tag = element_text(size = 7))

# Save plot
#ggsave(here("results", "figure_6.png"), width = 6.89, height = 2.5)
#ggsave(here("results", "figure_6.tiff"), width = 6.89, height = 2.5)

```

### Supplemental Figure

```{r plot_sup_freq_all, include=FALSE}
fractions_pairs <- G_fractions %>% 
    group_by(strain_pair, syncom, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3))

freq_plots <- function(name){
    plt <- fractions_pairs %>% 
        filter(strain_pair == name) %>% 
        ggplot(aes(x=r, y = mean_fraction))+
        facet_nested(synID + syncom ~ dpi, 
                     nest_line = element_line(color="black"),
                     labeller = labeller(dpi=dpi.lab2))+
        geom_area(aes(fill=fct_reorder(type, index)), color = "black", linewidth = 0.25)+
        scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                           expand = c(0,0), limit=c(0.2,30), breaks = seq(10, 20, 10))+
        scale_y_continuous(name = bquote("Relative frequency of"~hat(g)(italic(r))),
                           expand = c(0,0), breaks = seq(0,1,0.5))+
        scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
        theme_rs()+
        labs(title = pair.lab[name])+
        theme(panel.spacing.x = grid::unit(0.25, "line"),
              panel.spacing.y = grid::unit(0.5, "line"),
              strip.text = element_text(angle=0, vjust = 0.5, hjust = 0, 
                                          margin = margin(r = 5, l = 2), face = "plain"))
    return (plt)
}

list_plot <- unique_pair %>% 
    map(~freq_plots(.x))

```

```{r figure_s5, warning=FALSE, fig.id=TRUE, fig.cap="Frequency plots g", fig.width=6.89, fig.height=5, out.width="75%"}
wrap_plots(list_plot, ncol = 5, guides = "collect")+
    plot_annotation(tag_levels = "A") &
    theme(legend.position = "bottom",
          title = element_text(size = 6),
          axis.text = element_text(size = 5),
          axis.title = element_text(size = 6),
          strip.text = element_text(size = 6),
          plot.tag = element_text(size = 7),
          plot.margin = margin(1,0,1,0))

# Save plot
#ggsave(here("results", "figure_s5.png"), width = 6.89, height = 5)
#ggsave(here("results", "figure_s5.tiff"), width = 6.89, height = 5)

```
