---
title: "Resubmission"
author: "Rudolf Schlechter"
date: "2024-10-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

# Libraries
library(here)
library(tidyverse)
library(rstatix)
library(compositions)

# Plotting 
library(ggprism)
library(patchwork)
library(grid)

# Other dependencies
source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))
source(here('code', 'function_fold_change.R'))
source(here('code', 'function_triad.R'))

set.seed(19900725)

```

```{r}
# Load data
data_cfu <- read.csv(here("results", "data_processed.csv")) %>% 
    na.omit %>% 
    mutate(dpi = factor(dpi),
           synID = factor(synID),
           taxa = factor(taxa))
```


####  Changes in taxon-specific population density correlate with community complexity ####

# Effect of SynCom type

Influence of community complexity on individual bacterial populations at the CFU level:

```{r}
## Statistical tests
data_cfu %>% 
    kruskal_test(cfu_log ~ synID)

data_cfu %>% 
    group_by(dpi) %>% 
    kruskal_test(cfu_log ~ synID)

data_cfu %>% 
    dunn_test(cfu_log ~ synID, p.adjust.method = "holm")

## Fold change
fold_change(data_cfu,
            value = "cfu",
            ref_variable = "synID",
            ref_group = "C")
```

This effect was reflected by a significant two-fold increase in S2 and five-fold decrease in S3


# Effect of sampling time

```{r}
## Statistical tests
data_cfu %>% 
    wilcox_test(cfu_log ~ dpi)

## Fold change
fold_change(data_cfu,
            value = "cfu",
            ref_variable = "dpi",
            ref_group = "07dpi")

fold_change(data_cfu,
            value = "cfu",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = "dpi")
```

# Effect of taxa

```{r}
## Statistical tests
data_cfu %>% 
    group_by(dpi, synID) %>% 
    wilcox_test(cfu_log ~ taxa)

## Fold change
fold_change(data_cfu,
            value = "cfu",
            ref_variable = "taxa",
            ref_group = "Methylobacterium",
            group_by_variable = c("dpi", "synID"))

#   Data frame for plotting
fc_cfu_taxa <- fold_change(data_cfu,
            value = "cfu",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = c("dpi", "strain", "taxa"),
            additional_variable = c("exp", "sample"))

```

# Effect of strain
```{r}
## Statistical tests
data_cfu %>% 
    group_by(dpi, strain) %>% 
    kruskal_test(cfu_log ~ synID)

data_cfu %>% 
    group_by(dpi, strain) %>% 
    dunn_test(cfu_log ~ synID, p.adjust.method = "holm") %>% 
    print(n = 30)

## Fold change
fc_cfu_strain <- fold_change(data_cfu,
            value = "cfu",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = c("dpi", "strain"),
            additional_variable = c("exp", "sample"))

```

## Plot 2
```{r}
source(here('code', 'plot', 'figure2.R'))
```


### Relative and total abundance data

```{r}
## Function to process data for calculating CLR from relative abundance data 
calculate_clr <- function(df){
    pivot_wider(df, 
                id_cols = strain, 
                names_from = dpi, 
                values_from = relative_fraction) %>% 
        column_to_rownames(var = "strain") %>% 
        mutate(across(ends_with("dpi"), compositions::clr)) %>% 
        mutate(across(ends_with("dpi"), as.vector))
}
```


```{r}
# Split data set between DPI and SYNCOMs
datatest <- split(data_cfu, list(data_cfu$dpi, data_cfu$syncom))

# Calculate relative fractions
relative_fractions_list <- lapply(datatest, function(df) {
    df %>%
        group_by(dpi, synID, syncom, taxa, strain) %>% 
        summarise(cfu = median(cfu), .groups = "drop") %>% 
        mutate(relative_fraction = cfu / sum(cfu))
    })

# Combine lists into a data frame
relative_fractions <- do.call(rbind, relative_fractions_list) %>% 
    filter(synID != "C")

# Transform data to centered log ratio
clr_relfraction <- relative_fractions %>% 
    split(., .$syncom) %>% 
    lapply(., calculate_clr)

#   Apply t-test to CLR data
t_test_results <- clr_relfraction %>% 
    lapply(., function(df) {
        t.test(df[[1]], df[[2]])
        })

# Extract p-values from the t.test() results
p_values <- lapply(t_test_results, function(result) result$p.value) %>% do.call(rbind, .)

# Extract t-statistics from the t.test() results
t_statistics <- lapply(t_test_results, function(result) result$statistic) %>% do.call(rbind, .)

# View the extracted p-values and t-statistics
cbind(p_values, t_statistics)

```

```{r}
# Summary statistics
relative_fractions %>% 
    group_by(strain) %>% 
    mutate(percentage = 100*relative_fraction) %>% 
    summarise(median = sprintf("%1.0f", median(percentage)),
              IQ1 = sprintf("%1.0f", quantile(percentage, probs = 0.25)),
              IQ3 = sprintf("%1.0f", quantile(percentage, probs = 0.75)),
              median_IQR = paste0(median,"% [IQR: ",IQ1,"-",IQ3,"%]"))
```


## Plot S2
```{r}
source(here('code', 'plot', 'figureS2.R'))
```

##### TRITAGONISTS #####

First, we create a data set in which we can compare the species pairs and the tritagonist

```{r}
# Split the data
split_data <- data_cfu %>% 
    dplyr::select(synID, syncom, strain, taxa) %>% 
    unique() %>% 
    split(., 
          list(.$synID,
               .$syncom),
          drop = TRUE,
          sep = "_")

# Apply the function to each group
# The pair_combinations function will create a data frame
# with the strain pairs and tritagonists
result_list <- lapply(split_data, pair_combinations)

# Combine the results into a single dataframe
final_result <- do.call(rbind, result_list)

# Clean data to create a complete data frame with pairs of taxa and 
# tritagonists
strain_pairs <- final_result %>% 
    rownames_to_column(var = "label") %>% 
    separate(label, 
             into = c("synID", "syncom"), 
             sep = "_")  %>% 
    separate(syncom, into = c("syncom", "part1"), sep = "\\.") %>% 
    unite(syncom:part1, 
          col = "syncom",
          sep = ".") %>% 
    unite(strain1:strain2, 
          col = "pair", 
          remove = FALSE) %>% 
    mutate(taxa_pair = case_when(
            pair == "meL85_meL92" | 
                pair == "meL85_meL92" | 
                pair == "meL85_mr01" | 
                pair == "meL92_mr01" ~ "MM",
            pair == "meL85_smfr1" | 
                pair == "meL85_spfa2" | 
                pair == "meL92_smfr1" | 
                pair == "meL92_spfa2" | 
                pair == "mr01_smfr1" | 
                pair == "mr01_spfa2" ~ "MS",
            pair == "smfr1_spfa2" ~ "SS"),
        taxa_tritagonist = case_when(
            tritagonist == "smfr1" | 
                tritagonist == "spfa2" ~ "Sphingomonas",
            tritagonist == "meL85" | 
                tritagonist == "meL92" | 
                tritagonist == "mr01" ~ "Methylobacterium")) %>% 
    mutate(taxa_tritagonist = factor(taxa_tritagonist),
           taxa_pair = factor(taxa_pair)) %>% 
    filter(synID != "C")

## Long version
strain_pairs_long <- strain_pairs %>% 
    pivot_longer(cols = strain1:strain2, 
                 names_to = "type_strain", 
                 values_to = "strain") %>% 
    dplyr::select(-type_strain)

```

```{r}
# Create a data frame from the CFU data with tritagonists, pairs of
# species and their identities. This way, we compare the CFU of individual
# strains that belong to a pair and calculate the fold change in relation
# to the pair in S2
data_cfu_full <- data_cfu %>% 
    filter(synID != "C") %>% 
    inner_join(., strain_pairs_long, 
               by = join_by(synID, syncom, strain),
               relationship = "many-to-many")

# Define de variables to group the data frame for fold change calculation
variables <- c("dpi", "pair", "strain", "taxa_pair")

## Fold change
strain_pairs_simple <- strain_pairs_long %>% 
    filter(synID == "S3")

reference_S2 <- data_cfu_full %>% 
    filter(synID == "S2") %>% 
    group_by(dpi, strain, taxa, pair, taxa_pair) %>% 
    summarise(reference_value = median(cfu))

fold_change_S3 <- data_cfu_full %>% 
    filter(synID == "S3") %>% 
    dplyr::select(exp, sample, dpi, 
                  synID, syncom, strain, 
                  taxa, cfu, pair, tritagonist, 
                  taxa_pair, taxa_tritagonist) %>% 
    inner_join(., reference_S2) %>% 
    mutate(FC = cfu/reference_value,
           log2FC = log2(FC))

```

Q: Is the fold change of Sphingomonas and Methylobacterium within a pair different in S3?

```{r}
##  One-sample Wilcoxon
##  Evaluate if FC of a strain in S3 is different to S2
fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    wilcox_test(log2FC ~ 1, mu = 0)

##  Wilcoxon test
##  Evaluate differences between Sphingomonas and Methylobacterium
fold_change_S3 %>% 
    group_by(dpi) %>% 
    wilcox_test(log2FC ~ taxa)

##  Fold changes of each taxa in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))
```

Q: Is the fold change of individual strains within a pair different in S3?

```{r}
##  One-sample Wilcoxon
##  Evaluate if FC of a strain in S3 is different to S2
fold_change_S3 %>% 
    group_by(dpi, strain) %>% 
    wilcox_test(log2FC ~ 1, mu = 0)

##  Kruskal-Wallis test
##  Evaluate differences between strains
fold_change_S3 %>% 
    group_by(dpi) %>% 
    kruskal_test(log2FC ~ strain)

##  Post-hoc test
fold_change_S3 %>% 
    group_by(dpi) %>% 
    dunn_test(log2FC ~ strain, p.adjust.method = "holm")

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, strain) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))
```

Q: Do these changes depend on the third species?

```{r}
fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    wilcox_test(log2FC ~ taxa_tritagonist)

fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    dunn_test(log2FC ~ tritagonist, p.adjust.method = "holm")

```

Q: Are taxa pairs different?

```{r}
fold_change_S3 %>% 
    group_by(dpi) %>% 
    kruskal_test(log2FC ~ taxa_pair)

fold_change_S3 %>% 
    group_by(dpi) %>% 
    dunn_test(log2FC ~ taxa_pair, p.adjust.method = "holm")

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa_pair) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))
```

Q: Do these differences depend on the tritagonist?

```{r}
fold_change_S3 %>% 
    group_by(dpi, taxa_pair) %>% 
    kruskal_test(log2FC ~ tritagonist)

fold_change_S3 %>% 
    group_by(dpi, taxa_pair) %>% 
    dunn_test(log2FC ~ tritagonist, p.adjust.method = "holm") %>% 
    filter(p.adj < 0.05)

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa_pair, tritagonist) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC)) %>% 
    arrange(-fold_decrease)
```

## Plots 3 and S3
```{r}
source(here('code', 'plot', 'figure3.R'))
```


