---
title: "Resubmission Microscopy data"
author: "Rudolf Schlechter"
date: "2024-10-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

# Libraries
## Data processing
library(here)
library(tidyverse)

## Stats
library(rstatix)
library(car)

## Data visualisation
library(ggdist)
library(ggpubr)
library(ggprism)
library(patchwork)
library(grid)

# Dependencies
source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))
source(here('code', 'function_triad.R'))

# Set seed
set.seed(19900725)
```

```{r}
# Load data
data_cell <- read.csv(here("results", "cell_density.csv")) %>% 
    inner_join(., read.csv(here('data', 'comm_id.csv'))) %>% 
    inner_join(., unique(read.csv(here("results", "data_processed.csv"))[,c(7,8)]), 
        by = "strain")
```

#### Effect of community complexity on intraspecific spatial relations ####

# Effect of SynCom type

Influence of community complexity on individual bacterial populations at the CFU level:

```{r}
## Statistical tests
data_cell %>% 
    kruskal_test(logCell ~ synID)

data_cell %>% 
    group_by(dpi) %>% 
    kruskal_test(logCell ~ synID)

data_cell %>% 
    dunn_test(logCell ~ synID, p.adjust.method = "holm")

## Fold change
fold_change(data_cell,
            value = "cell_density",
            ref_variable = "synID",
            ref_group = "C") %>% 
    mutate(invFC = 1/FC)
```

This effect was reflected by a significant two-fold increase in S2 and five-fold decrease in S3

# Effect of sampling time

```{r}
## Statistical tests
data_cell %>% 
    wilcox_test(logCell ~ dpi)
```

# Effect of taxa

```{r}
## Statistical tests
data_cell %>% 
    group_by(dpi, synID) %>% 
    wilcox_test(logCell ~ taxa)

## Fold change
fold_change(data_cell,
            value = "cell_density",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = c("dpi", "taxa")) %>% 
    mutate(invFC = 1/FC)

#   Data frame for plotting
fc_cell_taxa <- fold_change(data_cell,
            value = "cell_density",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = c("dpi", "strain", "taxa"),
            additional_variable = c("exp", "rep"))

```

# Effect of strain
```{r}
## Statistical tests
data_cell %>% 
    group_by(dpi, strain) %>% 
    kruskal_test(logCell ~ synID)

data_cell %>% 
    group_by(dpi, strain) %>% 
    dunn_test(logCell ~ synID, p.adjust.method = "holm") %>% 
    print(n = 30)

fold_change(data_cell,
            value = "cell_density",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = c("dpi", "strain")) %>% 
    mutate(invFC = 1/FC) %>% 
    arrange(log2FC)

## Fold change
fc_cell_strain <- fold_change(data_cell,
            value = "cell_density",
            ref_variable = "synID",
            ref_group = "C",
            group_by_variable = c("dpi", "strain"),
            additional_variable = c("exp", "rep"))

```

## Plot 5
```{r}
source(here('code', 'plot', 'figure5.R'))
```

##### TRITAGONISTS #####

First, we create a data set in which we can compare the species pairs and the tritagonist

```{r}
# Split the data
split_data <- data_cell %>% 
    dplyr::select(synID, syncom, strain, taxa) %>% 
    unique() %>% 
    split(., 
          list(.$synID,
               .$syncom),
          drop = TRUE,
          sep = "_")

# Apply the function to each group
# The pair_combinations function will create a data frame
# with the strain pairs and tritagonists
result_list <- lapply(split_data, pair_combinations)

# Combine the results into a single dataframe
final_result <- do.call(rbind, result_list)

# Clean data to create a complete data frame with pairs of taxa and 
# tritagonists
strain_pairs <- final_result %>% 
    rownames_to_column(var = "label") %>% 
    separate(label, 
             into = c("synID", "syncom"), 
             sep = "_")  %>% 
    separate(syncom, into = c("syncom", "part1"), sep = "\\.") %>% 
    unite(syncom:part1, 
          col = "syncom",
          sep = ".") %>% 
    unite(strain1:strain2, 
          col = "pair", 
          remove = FALSE) %>% 
    mutate(taxa_pair = case_when(
            pair == "meL85_meL92" | 
                pair == "meL85_meL92" | 
                pair == "meL85_mr01" | 
                pair == "meL92_mr01" ~ "MM",
            pair == "meL85_smfr1" | 
                pair == "meL85_spfa2" | 
                pair == "meL92_smfr1" | 
                pair == "meL92_spfa2" | 
                pair == "mr01_smfr1" | 
                pair == "mr01_spfa2" ~ "MS",
            pair == "smfr1_spfa2" ~ "SS"),
        taxa_tritagonist = case_when(
            tritagonist == "smfr1" | 
                tritagonist == "spfa2" ~ "Sphingomonas",
            tritagonist == "meL85" | 
                tritagonist == "meL92" | 
                tritagonist == "mr01" ~ "Methylobacterium")) %>% 
    mutate(taxa_tritagonist = factor(taxa_tritagonist),
           taxa_pair = factor(taxa_pair)) %>% 
    filter(synID != "C")

## Long version
strain_pairs_long <- strain_pairs %>% 
    pivot_longer(cols = strain1:strain2, 
                 names_to = "type_strain", 
                 values_to = "strain") %>% 
    dplyr::select(-type_strain)

```

```{r}
# Create a data frame from the microscopy data with tritagonists, pairs of
# species and their identities. This way, we compare the cell numbers of individual
# strains that belong to a pair and calculate the fold change in relation
# to the pair in S2
data_cell_full <- data_cell %>% 
    filter(synID != "C") %>% 
    inner_join(., strain_pairs_long, 
               by = join_by(synID, syncom, strain),
               relationship = "many-to-many")

# Define de variables to group the data frame for fold change calculation
variables <- c("dpi", "pair", "strain", "taxa_pair")

## Fold change
strain_pairs_simple <- strain_pairs_long %>% 
    filter(synID == "S3")

reference_S2 <- data_cell_full %>% 
    filter(synID == "S2") %>% 
    group_by(dpi, strain, taxa, pair, taxa_pair) %>% 
    summarise(reference_value = median(cell_density))

fold_change_S3 <- data_cell_full %>% 
    filter(synID == "S3") %>% 
    dplyr::select(exp, rep, dpi, 
                  synID, syncom, strain, 
                  taxa, cell_density, pair, tritagonist, 
                  taxa_pair, taxa_tritagonist) %>% 
    inner_join(., reference_S2) %>% 
    mutate(FC = cell_density/reference_value,
           log2FC = log2(FC))

```

Q: Is the fold change of Sphingomonas and Methylobacterium within a pair different in S3?

```{r}
##  One-sample Wilcoxon
##  Evaluate if FC of a strain in S3 is different to S2
fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    wilcox_test(log2FC ~ 1, mu = 0) # All different except for Sphingomonas at 14 dpi are not different to S2

##  Wilcoxon test
##  Evaluate differences between Sphingomonas and Methylobacterium
fold_change_S3 %>% 
    group_by(dpi) %>% 
    wilcox_test(log2FC ~ taxa) # Taxa are different in both time points

##  Fold changes of each taxa in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))
```

Q: Is the fold change of individual strains within a pair different in S3?

```{r}
##  One-sample Wilcoxon
##  Evaluate if FC of a strain in S3 is different to S2
fold_change_S3 %>% 
    group_by(dpi, strain) %>% 
    wilcox_test(log2FC ~ 1, mu = 0)

##  Kruskal-Wallis test
##  Evaluate differences between strains
fold_change_S3 %>% 
    group_by(dpi) %>% 
    kruskal_test(log2FC ~ strain)

##  Post-hoc test
fold_change_S3 %>% 
    group_by(dpi) %>% 
    dunn_test(log2FC ~ strain, p.adjust.method = "holm")

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, strain) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))
```

Q: Do these changes depend on the third species?

```{r}
## Taxa-level tritagonist
fold_change_S3 %>% 
    group_by(dpi, taxa) %>% 
    wilcox_test(log2FC ~ taxa_tritagonist)

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa, taxa_tritagonist) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))

##  Strain-level tritagonist
fold_change_S3 %>% 
    filter(taxa == "Methylobacterium") %>% 
    group_by(dpi, tritagonist, synID) %>% 
    summarise(log2FC = median(log2FC)) %>% 
    filter(tritagonist %in% c("meL85", "meL92", "mr01"))

```

Q: Are taxa pairs different?

```{r}
fold_change_S3 %>% 
    group_by(dpi) %>% 
    kruskal_test(log2FC ~ taxa_pair)

fold_change_S3 %>% 
    group_by(dpi) %>% 
    dunn_test(log2FC ~ taxa_pair, p.adjust.method = "holm")

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa_pair) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC))
```

Q: Do these differences depend on the tritagonist?

```{r}
fold_change_S3 %>% 
    group_by(dpi, taxa_pair) %>% 
    kruskal_test(log2FC ~ tritagonist)

fold_change_S3 %>% 
    group_by(dpi, taxa_pair) %>% 
    dunn_test(log2FC ~ tritagonist, p.adjust.method = "holm") %>% 
    filter(p.adj < 0.05)

##  Fold changes of strains in S3 compared to S2
fold_change_S3 %>% 
    group_by(dpi, taxa_pair, tritagonist) %>% 
    summarise(median_log2FC = median(log2FC),
              fold_decrease = 2^abs(median_log2FC)) %>% 
    arrange(-fold_decrease)
```

## Plot S5
## Effect of tritagonists on fold change population cell density (microscopy data)
```{r}
source(here('code', 'plot', 'figureS5.R'))
```
