---
title: "Bacterial community complexity in the phyllosphere penalises specialists over generalists - Results Section 2"
author: "Rudolf Schlechter"
output:
  html_document:
    df_print: paged
    keep_md: yes
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

# Libraries
## Data processing
library(here)
library(tidyverse)

## Stats
library(rstatix)
library(car)

## Data visualisation
library(ggdist)
library(ggpubr)
library(patchwork)

# Dependencies
source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))
source(here('code', 'dunnet_function.R'))

# Set seed
set.seed(19900725)
```

## Spatial distribution of individual strains depend on their community context

We obtained an exemplary representation of every bacterial population at the single cell resolution, as well as their spatial distributions. This was accomplished by determining the centre of mass of individual cells within each bacterial combination. The centres of mass of the different cell populations were then used to generate spatial point patterns and further used to analyse spatial distribution patterns of bacteria. In total, we analysed 10,261 fields of view, corresponding to 25 distinct communities (5× C, 10× S2, 10× S3), captured at two time points, encompassing 3 to 4 biological replicates (individual plants). On average, each biological replicate was composed of 55 ± 30 fields of view (mean ± SD), in two independent experiments. 

```{r}
# Load spatial test for homogeneity
# Spatial Kolmogorov-Smirnov Test
cdf <- readRDS(here('results', 'cdf_fit1.rds')); cdf

# Chi-square quadrat test with Monte Carlo simulations
quadrat_test <- readRDS(here('results', 'quadrat_test_fit1.rds')); quadrat_test
```

We identified several characteristic patterns that follow the leaf surface topography, including veins and epidermal cell grooves, each exhibiting varying levels of occupancy. Subsequently, we observed that the cell distribution on the leaf surface was heterogenous by subsampling the data set to 500 random fields of view. We confirmed that these patterns can be described as a non-homogeneous Poisson process by a spatial Kolmogorov-Smirnov test (*D* = `r cdf$statistic`, *p* = `r cdf$p.value`), and a goodness-of-fit with Monte Carlo test ($X^{2}$ = `r quadrat_test$statistic`, *p* = `r quadrat_test$p.value`, n = 999).


```{r import_data, include=FALSE}
# Import and combine CFU and cell data
data_cell <- read.csv(here("results", "cell_density.csv")) %>% 
    rename(sample = rep) %>% 
    inner_join(., read.csv(here("results", "data_processed.csv")), 
               by=c("exp", "dpi", "synID", "comID", "syncom", "strain", "sample", "channel")) %>% 
    mutate(dpi = factor(dpi),
           synID = factor(synID),
           strain = factor(strain)) %>% 
    na.omit

# Correlation between bacterial densities determined by CFU or by cell counts
cell_corr <- data_cell %>% 
    cor_test(
        vars = c("logCell", "cfu_log"),
        method = "pearson",
        conf.level = 0.95)
```


Next, we estimated cell density per square centimetre for each biological replicate and compared our estimations of cell density from single-cell microscopy data [# cell cm^-2^] with colony count data [CFU gFW^-1^], which were positively correlated (Pearson's *r* = `r cell_corr$cor`, *t* = `r cell_corr$statistic`, *p* = `r cell_corr$p`).


```{r test_assumptions}
# Linear model to test normality and homogeneity of variance
linear_cell <-  lm(logCell ~ synID + dpi + taxa, data_cell)

# Shapiro-Wilk test for normality
shapiro.test(rstandard(linear_cell))

# Breusch-Pagan test for homogeneity of variances
ncvTest(linear_cell)
```

```{r cell_analysis_synID}
# Kruskal-Wallis test and effect size for community complexity (synID) grouped by taxa
kw_synID <- data_cell %>%
    group_by(taxa) %>% 
    kruskal_test(logCell ~ synID) %>% 
    mutate(p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))
keff_synID <- data_cell %>% 
    group_by(taxa) %>% 
    kruskal_effsize(logCell ~ synID, ci=TRUE, nboot=100)

## Variables
explanatory = "synID"
group = "taxa"
strain = "strain"

# Fold change of population density by taxa and synID
dunntest_taxa <- dun_func1(data_cell, "logCell", group, explanatory) %>% 
    mutate(symbol = ifelse(p_label == "< 0.05", "*", "")) %>% 
    rename(synID = group2)
fold_taxa <- fold_func1(data_cell, "cell_density", group, explanatory)

# Fold change of population density by strain and synID
dunntest_strain <- dun_func1(data_cell, "logCell", strain, explanatory) %>% 
    mutate(symbol = ifelse(p_label == "< 0.05", "*", "")) %>% 
    rename(synID = group2)
fold_strain <- fold_func1(data_cell, "cell_density", strain, explanatory)

```

```{r cell_analysis_dpi}
# Wilcoxon test and effect size for sampling time (dpi)
w_dpi = data_cell %>% 
    wilcox_test(formula = logCell ~ dpi)

# Fold change of population density by time of sampling (dpi)
fc_cell_dpi = data_cell %>% 
    group_by(dpi) %>% 
    summarise(median_cell = median(cell_density)) %>% 
    mutate(FC = median_cell/median_cell[1],
           logFC = log2(FC))
```

We observed a decrease in cell density between populations from S2 to S3 (Figure 4A). Differences in cell density were unrelated to time of sampling (Wilcoxon,  *W* = `r w_dpi$statistic`, *p* = `r w_dpi$p`), but was rather driven by a decrease in *Methylobacterium* populations (log~2~FC = `r fold_taxa$log2FC[fold_taxa$taxa=="Methylobacterium"&fold_taxa$synID=="S3"]`) and a marginal increase in *Sphingomonas* populations (log~2~FC = `r fold_taxa$log2FC[fold_taxa$taxa=="Sphingomonas"&fold_taxa$synID=="S3"]`). These differences were associated with a decline in MeL85 populations and an increase in SmFR1 within S3. These observations indicate that the effect of the community complexity in a population was consistent at the CFU level and the single-cell resolution.


## Figures

```{r figure_main_5_setup, include = FALSE}
# Plot CFU
# Multiple comparisons
plt.cell.stat = dunntest_taxa %>% 
    ungroup %>% 
    rename(group2 = synID) %>% 
    mutate(p.adj.label = case_when(p.adj < 0.05 ~ "*", TRUE ~ "ns"), 
           y.position = c(7.5,8.5,8,7.5,8.5,8))

plt.a <- data_cell %>% 
    ggplot(aes(synID, logCell))+
    facet_wrap(~taxa)+
    stat_eye(
        side="right",
        adjust = 1,
        justification = -0.3,
        .width = 0,
        scale = 0.5,
        point_colour = NA)+
    geom_jitter(width = 0.1, alpha = 0.1)+
    geom_boxplot(fill="white", width=0.2, outlier.alpha = 0)+
    coord_cartesian(xlim=c(1,3.2))+
    stat_pvalue_manual(
        data = plt.cell.stat,
        size = 5,
        label = "p.adj.label",
        xmin = "group1", xmax = "group2",
        y.position = "y.position")+
    scale_y_continuous(limits = c(2,9), breaks=c(4,6,8))+
    labs(y = plt_bac_cell_density_lab,
         x = "SynCom")+
    theme_rs()+
    theme(axis.text.x = element_text(hjust=0.5, vjust=1.5))

plt.b <- inner_join(dunntest_taxa, 
                     fold_taxa, 
                     by = c("taxa", "synID")) %>% 
    filter(group1 == "C") %>%
    na.omit %>% 
    ggplot(aes(x = synID, y = taxa))+
    geom_tile(aes(fill = log2FC), colour= "black")+
    geom_text(aes(label=symbol), vjust = 0.77, size = 5)+
    coord_equal(ratio = 0.5)+
    scale_fill_gradientn(name = bquote(Log[2]~"FC"), 
                         colours = wes_palette("Zissou1")[c(1,2,3,5)], 
                         values = c(0,0.67,1), 
                         limits=c(-3.5,1), 
                         breaks=seq(-3,1,1), 
                         na.value = 'grey90')+
    scale_y_discrete(name="", labels = taxa.lab)+
    labs(x = "")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust = 0.5, vjust = 3),
          axis.text.y = element_text(face="italic"),
          strip.text = element_text(face="plain"))

plt.c <- inner_join(dunntest_strain, fold_strain, by = c("strain", "synID")) %>% 
    filter(group1 == "C") %>% 
    ggplot(aes(x=synID, y=strain))+
    geom_tile(aes(fill = log2FC), colour= "black")+
    geom_text(aes(label=symbol), vjust = 0.77, size = 5)+
    coord_equal(ratio = 0.5)+
    scale_fill_gradientn(name = bquote(Log[2]~"FC"), 
                         colours = wes_palette("Zissou1")[c(1,2,3,5)], 
                         values = c(0,0.67,1), 
                         limits=c(-3.5,1), 
                         breaks=seq(-3,1,1), 
                         na.value = 'grey90')+
    scale_y_discrete(name="Focal strain", labels = sp.lab)+
    scale_x_discrete(name="SynCom", labels = sp.lab)+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust = 0.5, vjust = 3),
          strip.text = element_text(face="plain"))

```

```{r figure_4, fig.id=TRUE, fig.cap="Bacterial cell density in the arabidopsis phyllosphere"}

# Define plotting area
areas <- c(patchwork::area(1,1,3), # A
           patchwork::area(1,2,1), # B
           patchwork::area(2,2,3), # C
           patchwork::area(4,2,4)) # Legend

# Set up plot
wrap_elements(plot = plt.a) + plt.b + plt.c + guide_area() +
    plot_annotation(tag_levels = "A") + 
    plot_layout(design = areas, 
                guides = "collect",
                widths = c(3, 1)) & 
    theme(legend.box.just = "left",
          legend.position = "bottom",
          panel.spacing.x = unit(0.5, "lines"),
          plot.margin = margin(1,5,1,5)) &
    guides(fill = guide_colorbar(title.position="top", title.hjust = 0.5))

# Save plot
ggsave(here("results", "figure_4.png"), width = 7, height = 5)
ggsave(here("results", "figure_4.tiff"), width = 7, height = 5)

```
