---
title: 'Spatial distribution - Section 3'
author: "Rudolf Schlechter"
output: 
    html_document:
        df_print: paged
        keep_md: yes
    pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

library(here)
source(here('code', 'libraries_syncom.R'))
source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))
library(rstatix)

#Data
source(here('code', 'auc_fractions.R'))

set.seed(19900725)
```

## Effect of community complexity on intraspecific spatial relations

```{r}
fractions %>% head
```
To evaluate the effect of community context on spatial distributions within bacterial populations (intraspecific interactions), we first determined relative frequencies of a spatial pattern (aggregation, random, regular) based on the K(r) estimates for every condition. We then determined the area under the curve of the aggregation fraction of frequency plots and calculated the fractional change of aggregation compared to the near-isogenic control condition, C (Fig 6a).

```{r}
## Taxa
wilcox_one_taxa = auc_fold_change %>% 
    group_by(synID, dpi, type, taxa) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(synID, dpi, type, taxa, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

## Strain
wilcox_one_strain = auc_fold_change %>% 
    group_by(synID, dpi, type, strain) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(synID, dpi, type, strain, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))
```

```{r plot_sup}
wilcox_one_taxa %>% 
    filter(type == "aggregate_fraction") %>% 
    ggplot(aes(synID, taxa))+
    facet_wrap(~dpi, ncol = 2, labeller = labeller(dpi=dpi.lab))+
    geom_tile(color = "black", fill = "white", linewidth = 0.1)+
    geom_point(aes(fill = estimate, size = p_size), shape = 21)+
    coord_fixed()+
    scale_fill_gradientn(name = "Change in\naggregation",
        colours = wes_palette("Zissou1")[c(1,2,3,5)], values=c(0,0.55,1), 
        limits = c(-1,0), breaks = seq(-1,0, 0.2))+
    scale_size_continuous(name = expression(paste(italic("P"), "-value")), 
                          range = c(12,2), breaks = c(0.05, 0.5, 1), limits = c(0,1),
                          labels = c("< 0.05", "0.5", "1.0"))+
    labs(x="", y="")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust=0.5, vjust=3),
          axis.text.y = element_text(face="italic"),
          strip.text = element_text(face="plain"))

wilcox_one_strain %>% 
    filter(type == "aggregate_fraction") %>% 
    ggplot(aes(synID, strain))+
    facet_wrap(~dpi, ncol = 2, labeller = labeller(dpi=dpi.lab))+
    geom_tile(color = "black", fill = "white", linewidth = 0.1)+
    geom_point(aes(fill = estimate, size = p_size), shape = 21)+
    coord_fixed()+
    scale_fill_gradientn(name = "Change in\naggregation",
        colours = wes_palette("Zissou1")[c(1,2,3,5)], values=c(0,0.55,1), 
        limits = c(-1,0), breaks = seq(-1,0, 0.2))+
    scale_size_continuous(name = expression(paste(italic("P"), "-value")), 
                          range = c(12,2), breaks = c(0.05, 0.5, 1), limits = c(0,1),
                          labels = c("< 0.05", "0.5", "1.0"))+
    scale_y_discrete(name = "Focal strain", labels = sp.lab)+
    labs(x="")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust=0.5, vjust=3),
          strip.text = element_text(face="plain"))
    
```


We first evaluated whether the spatial pattern of a population in a community was different from its near-isogenic control (C), using a one-sample Wilcoxon test. The null hypothesis was that there is no change of a spatial pattern within a population in S2 or S3. On average, we observed a 23.9% increase in aggregation of Methylobacterium strains (t17 = 6.58, p < 0.05) at 14 dpi in S3 (Table SX). This increase was present in every Methylobacterium strain (Fig. 6b, MeL85 = +28.6%; MeL92 = +16.6%; Mr0-1 = +26.6%). Within the sphingomonads, SmFR1 decreased its aggregation pattern by 11.6% at 7 dpi in S3 communities (t5 = -5.03, p < 0.05), while SpFA2 remained unchanged (Fig 6b). 

We defined the maximal intraspecific aggregation distances of a population as the maximal distance in which K(r) is higher than the upper limit of the estimator Kinhom(r) envelope, which indicates spatial aggregation (Fig 2b, Fig 6a). We determined the intraspecific aggregation distances for each interaction until they reached the maximal aggregation distance using the empirical cumulative distribution function (Fig. 6a), in which 95% of the aggregation was observed up to a given distance. We observed differences in intraspecific aggregation distances between strains: MeL85, MeL92, and SpFA2 showed the largest aggregation ranges of, in a few cases, up to 35 μm, while Mr0-1 and SmFR1 showed the shorter aggregation ranges of 0–15.5 µm and 0–18 μm, respectively (Fig. 6c). These ranges were not explained by the treatments or sampling points, but only by strain differences (F4,81 = 6.53, p < 0.05).


```{r}
auc_fold_change %>% 
    ggplot(aes(syncom, strain))+
    facet_grid(dpi~type)+
    geom_tile(colour='black', fill='white')+
    geom_point(aes(fill=fractional_change), shape = 21, size = 6)+
    scale_fill_gradientn(colours = wes_palette("Zissou1")[c(1,3,5,5)], values = c(0,0.35,1))+
    coord_fixed()

auc_fold_change %>% 
    ggplot(aes(dpi, fractional_change))+
    facet_grid(type~strain)+
    geom_point()

auc_fold_change %>% 
    ggplot(aes(auc.inter))+
    geom_histogram()

lmauc = lm(fractional_change ~ synID + dpi + type + strain, data = auc_fold_change)
shapiro.test(rstandard(lmauc))
ncvTest(lmauc)


## synID
auc_fold_change %>% 
    filter(type == "aggregate_fraction") %>% 
    group_by(dpi) %>% 
    wilcox_test(fractional_change ~ synID, p.adjust.method = "holm")

## dpi
auc_fold_change %>% 
    filter(type == "aggregate_fraction") %>% 
    wilcox_test(fractional_change ~ dpi, detailed = TRUE)

## strain
auc_fold_change %>% 
    filter(type == "aggregate_fraction") %>% 
    group_by(dpi) %>% 
    kruskal_test(fractional_change ~ strain)

auc_fold_change %>% 
    filter(type == "aggregate_fraction") %>% 
    group_by(dpi) %>% 
    dunn_test(fractional_change ~ strain, p.adjust.method = "holm")

```

```{r plot_sup}
auc_fold_change %>% 
    ggplot(aes(taxa, fractional_change))+
    facet_grid(dpi ~ type, labeller = labeller(dpi = dpi.lab, type = pattern.lab))+
    geom_jitter(aes(color = strain), width = 0.1, alpha = 0.8)+
    geom_boxplot(alpha = 0.5, fill = "white", width = 0.2, outlier.alpha = 0)+
    geom_hline(yintercept = 0, linetype = "dashed")+
    theme_rs()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face="italic"),
          strip.text = element_text(face = "plain"))+
    labs(x = "", y = "Fractional change")+
    scale_color_manual(name = "Strain", values=sp.pal, labels=sp.lab)
```