---
title: 'Spatial distribution - Section 4'
author: "Rudolf Schlechter"
output: 
    html_document:
        df_print: paged
        keep_md: yes
    pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

library(here)
library(tidyverse)
source(here('code', 'libraries_syncom.R'))
source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))
library(rstatix)

#Data
source(here('code', 'auc_pcf_fractions.R'))
source(here('code', 'function_fractional_change_median.R'))
auc_fold_change <- auc_fold_change %>% 
    mutate(taxa_type = case_when(
        taxa_pair == "MM" | taxa_pair == "SS" ~ "within",
        taxa_pair == "MS" ~ "between"))
auc_aggregation <- auc_fold_change %>% filter(type == "aggregate_fraction")


set.seed(19900725)
```

### Effect of community complexity on interspecific spatial correlations

We investigated the effect of community complexity on interspecific interactions by determining the co-aggregation frequencies of pairs of species in different treatments (S2 and S3) and sampling times (7 and 14 dpi). Co-aggregation was determined through pair cross correlations at a given distance, g(r), between pairs of strains (Fig. 7a, Fig. S4). Frequencies were then determined from all replicates and the area under the curve was used as a measure to compare aggregation patterns of pairs of species in S3 relative to S2 communities.

```{r analysis_type}
## Summary
summary_type <- fractional_change_median(auc_fold_change, group = "type")

## Kruskal-Wallis
kw_type <- auc_fold_change %>% 
    kruskal_test(fractional_change ~ type) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

kw_eff_type <- auc_fold_change %>% 
    kruskal_effsize(fractional_change ~ type, ci=TRUE, nboot=100)

## One-sample Wilcoxon test
w1_type <-  auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

```


```{r analysis_dpi}
## Summary
summary_dpi <- fractional_change_median(auc_fold_change, group="dpi")

## One sample Wilcoxon test
w1_dpi = auc_fold_change %>% 
    group_by(dpi, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

## Two samples Wilcoxon test
w2_dpi <- auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ dpi, p.adjust.method = "holm", detailed = TRUE)
```

```{r analysis_taxa_strain}
## Two samples Wilcoxon test for taxa type
w2_taxaType <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ taxa_type, p.adjust.method = "holm", detailed = TRUE)

## One sample Wilcoxon test for taxa type
w1_taxaType <-  auc_fold_change %>% 
    group_by(dpi, type, taxa_type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, taxa_type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

## Two samples Wilcoxon test for taxa pairs
kw_taxa <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ taxa_pair)

dunn_taxa <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    dunn_test(fractional_change ~ taxa_pair, p.adjust.method = "holm", detailed = TRUE) %>% 
    filter(p.adj < 0.05)

## One sample Wilcoxon test for taxa pairs
w1_taxa <-  auc_fold_change %>% 
    group_by(dpi, type, taxa_pair) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, taxa_pair, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "*", TRUE ~ ""))

## Kruskal Wallis test for strain pairs
kw_strain <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ strain_pair)

dunn_strain <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    dunn_test(fractional_change ~ strain_pair, p.adjust.method = "holm", detailed = TRUE) %>% 
    select(dpi, type, group1, group2, p, p.adj, p.adj.signif)

## Two samples Wilcoxon test for strain pairs
w2_strain <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ strain_pair, p.adjust.method = "holm", detailed = TRUE)
w2_strain %>% filter(p.adj < 0.05)

## One sample Wilcoxon test for strain pairs
w1_strain  <-  auc_fold_change %>% 
    group_by(dpi, strain_pair, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, p.adjust.method = "holm", detailed = TRUE) %>% 
    select(dpi, strain_pair, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))
w1_strain %>% filter(p < 0.05)

```

```{r}
## Kruskal Wallis test for strain pairs
kw_syncom <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ syncom)

## Two samples Wilcoxon test for SynCom
w2_syncom <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ syncom, p.adjust.method = "holm", detailed = TRUE) %>% 
    filter(p.adj < 0.05)

## One sample Wilcoxon test for taxa pairs
w1_syncom <-  auc_fold_change %>% 
    group_by(dpi, type, syncom) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, syncom, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "*", TRUE ~ as.character(p)))
w1_syncom %>% filter(p_label == "*")
```


```{r, include = FALSE}
plt.frq <- fractions %>% 
    group_by(taxa_pair, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(x=r, y=mean_fraction))+
    facet_nested(taxa_pair ~ dpi + synID, 
                 nest_line = element_line(color="black"),
                 labeller = labeller(dpi = dpi.lab2))+
    geom_area(aes(fill=fct_reorder(type, index)), color = "black")+
    scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                       expand = c(0,0), limit=c(0.2,30), breaks = seq(5, 25, 10))+
    scale_y_continuous(name = expression(paste("Relative frequency of g(", italic(r),")")), 
                       expand = c(0,0), breaks = seq(0,1,0.5))+
    scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
    theme_rs()+
    theme(
          panel.spacing.x = grid::unit(0.5, "line"),
          strip.text.x = element_text(margin = margin(b = 1, t = 5), face = "plain"),
          strip.text.y = element_text(angle=0, vjust = 0.5, hjust = 0, face = "plain"),
          legend.position = "bottom")

plt.pattern <- left_join(auc_fold_change, w1_taxa, by = c("dpi", "type", "taxa_pair")) %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(taxa_pair, fractional_change))+
    geom_text(aes(y=2.5, label=p_label), size = 5, parse=FALSE)+
    facet_grid(dpi ~ fct_reorder(type, index), labeller = labeller(.rows = dpi.lab2, .cols = pattern.lab))+
    geom_jitter(color = "black", stroke = 0.5, width = 0.1, alpha = 0.8)+
    geom_boxplot(fill="white", alpha = 0.5, width = 0.2, outlier.alpha = 0)+
    geom_hline(yintercept = 0, linetype = "dashed")+
    theme_rs()+
    theme(strip.text = element_text(face = "plain"))+
    labs(x = "Taxa pair", y = "Fractional change")

```

```{r figure_main_6, fig.id=TRUE, dev = "png", fig.dim=c(5,8)}
plt.frq / wrap_elements(plot = plt.pattern) + 
    plot_annotation(tag_levels = "A") &
    theme(legend.box.just = "center",
          plot.margin = margin(0,1,0,1))
```


We noticed that the frequencies of co-aggregation were generally lower than those observed from within-population aggregation. However, co-aggregation was detected in every strain pair in S2 communities at each sampling point (Fig. S4). In general, the presence of a third strain decreased these co-aggregation frequencies. We observed that co-aggregation was 23% lower in S3 than S2 at 7 dpi (one sample t-test, t29 = -2.37, p = 0.024), but became comparable at 14 dpi (Fig. 7b, one sample t-test, t29 = 0.66, p = 0.51). Co-aggregation at 14 dpi was higher than at 7 dpi in S3 (two sample t-test, t58 = -2.09, p = 0.041).

We determined the maximal frequency of co-aggregation for each bacterial pair, which was generally observed at a ~1.5 μm distance between cells (Fig. 7a, Fig. S4). We observed an increase of co-aggregation over time, as maximal co-aggregation frequencies increased from 20.3 ± 12.6% to 40.8 ± 16.5% at 7 and 14 dpi, respectively (Fig. 7c). 

In general, we observed that co-aggregation frequencies decreased rapidly after 1.5 μm. Since co-aggregation was detected at every distance range (Fig. S5a), we focused on the differences in co-aggregation only within the 0–10 μm distance interval instead of the maximal range of co-aggregation. Based on the empirical cumulative distribution function (Fig. S5b), 64% of the co-aggregation pattern was observed at a distance equal or less than 10 μm, in which the maximal frequency of co-aggregation was captured. We observed that at 14 dpi, a high percentage of bacterial pairs (66.6%) co-aggregated up to 10 μm, in comparison to 53.1% at 7 dpi (Fig. 7d, F1,71 = 17.79, p < 0.05). Although no differences were observed between S2 and S3 (F1,71 = 2.57, p = 0.11), we noticed that the initial variation in co-aggregation distances in S3 was markedly higher compared to S2 after 7 days (Fig. 7e). After 14 days, the variation in co-aggregation distances in S3 decreased and was comparable to S2 treatments. 

Our results indicate that the low co-aggregation and increased randomness observed in more complex communities are only transitional, as from 7 to 14 dpi, co-aggregation between pairs in presence of a third strain are comparable to conditions where only these pairs were present in the phyllosphere.


```{r plot_sup_freq_all, include=FALSE}
fractions_pairs <- fractions %>% 
    group_by(strain_pair, syncom, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3))

freq_plots <- function(name){
    plt <- fractions_pairs %>% 
        filter(strain_pair == name) %>% 
        ggplot(aes(x=r, y=mean_fraction))+
        facet_nested(synID + syncom ~ dpi, 
                     nest_line = element_line(color="black"),
                     labeller = labeller(dpi=dpi.lab2))+
        geom_area(aes(fill=fct_reorder(type, index)), color = "black")+
        scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                           expand = c(0,0), limit=c(0.2,30), breaks = seq(10, 20, 10))+
        scale_y_continuous(name = expression(paste("Relative frequency of g(", italic(r),")")), 
                           expand = c(0,0), breaks = seq(0,1,0.5))+
        scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
        theme_rs()+
        labs(title = pair.lab[name])+
        theme(panel.spacing.x = grid::unit(0.5, "line"),
              panel.spacing.y = grid::unit(1, "line"),
              strip.text = element_text(angle=0, vjust = 0.5, hjust = 0, 
                                          margin = margin(r = 5, l = 2), face = "plain"))
    return (plt)
}

list_plot <- unique_pair %>% 
    map(~freq_plots(.x))

```

```{r figure_supplemental_6, warning=FALSE, fig.id=TRUE, fig.cap="Frequency plots g", dev = "png", fig.dim=c(16,8)}
wrap_plots(list_plot, ncol = 5, guides = "collect")+
    plot_annotation(tag_levels = "A")
```

