---
title: 'Spatial distribution - Section 4'
author: "Rudolf Schlechter"
output: 
    html_document:
        df_print: paged
        keep_md: yes
    pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

library(here)
library(tidyverse)
library(rstatix)

source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))

#Data
source(here('code', 'spat_pcf_fractions_auc.R'))
source(here('code', 'function_fractional_change_median.R'))
auc_fold_change <- auc_fold_change %>% 
    mutate(taxa_type = case_when(
        taxa_pair == "MM" | taxa_pair == "SS" ~ "within",
        taxa_pair == "MS" ~ "between"))
auc_aggregation <- auc_fold_change %>% filter(type == "aggregate_fraction")


set.seed(19900725)
```

### Effect of community complexity on interspecific spatial correlations

We investigated the effect of community complexity on the spatial patterns between pairs of species (interspecific interactions). Pair cross correlations, $\hat{g}$(r), indicated whether an aggregated, random, or regular pattern was detected at a given distance (Fig. XXa, Fig. SXX). Frequencies were then determined from all replicates and the area under the curve was used as a measure to compare a spatial pattern of pairs of species in S3 relative to S2.

```{r analysis_type}
## Summary
summary_type <- fractional_change_median(auc_fold_change, group = "type")

## Kruskal-Wallis
kw_type <- auc_fold_change %>% 
    kruskal_test(fractional_change ~ type) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

kw_eff_type <- auc_fold_change %>% 
    kruskal_effsize(fractional_change ~ type, ci=TRUE, nboot=100)

## Dunn's test
dunn_type <- auc_fold_change %>% 
    dunn_test(fractional_change ~ type, p.adjust.method = "holm")

## One-sample Wilcoxon test
w1_type <-  auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

```

In general terms, pairs of species aggregated in S2 communities at each sampling point (Fig. XXA, Fig. SXX). The presence of a third strain decreased these aggregation frequencies, and consequently, regularity and randomness also changed in S3 (pair + one extra species) compared to the pairs alone (Kruskal-Wallis, *H*(`r kw_type$df`) = `r kw_type$statistic`, *p* `r kw_type$p_label`). Aggregation and regularity decreased in `r summary_type$percentage[summary_type$type=="aggregate_fraction"]` and `r summary_type$percentage[summary_type$type=="regular_fraction"]`%, respectively, while randomness was `r summary_type$percentage[summary_type$type=="random_fraction"]`% higher.

```{r analysis_dpi}
## Summary
summary_dpi <- fractional_change_median(auc_fold_change, group=c("dpi", "type"))

## One sample Wilcoxon test
w1_dpi = auc_fold_change %>% 
    group_by(dpi, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))

## Two samples Wilcoxon test
w2_dpi <- auc_fold_change %>% 
    group_by(type) %>% 
    wilcox_test(fractional_change ~ dpi, p.adjust.method = "holm", detailed=TRUE)
```

To further dissect these differences, we grouped the species pairs based on their identity, as a pair was composed of either two methylobacteria (MM), a methylobacterium and a sphingomonad (MS), or two sphingomonads (SS). We noticed that randomness was high at 7 dpi for pairs belonging to an S3 community (One-sample Wilcox, *W* = `r w1_dpi$statistic[w1_dpi$type=="random_fraction" & w1_dpi$dpi=="07dpi"]`, *p* `r w1_dpi$p_label[w1_dpi$type=="random_fraction" & w1_dpi$dpi=="07dpi"]`), and this increase was associated with pairs of species containing at least one *Methylobacterium* species (MM and MS) (Fig. XXB). However, this randomness decreased over time and became comparable to S2 at 14 dpi. The decrease in aggregation was only observed in MM pairs at 14 dpi. By contrast, pairs composed of only *Sphingomonas* species (SS) were not different between S2 and S3.

Our results indicate that *Methylobacterium* species are more susceptible than *Sphingomonas* to changes in their spatial distribution depending on their community context. The pairs containing at least one *Methylobacterium* were randomly distributed in more complex communities initially.

```{r analysis_taxa_strain}
## One sample Wilcoxon test for taxa pairs
w1_taxa <-  auc_fold_change %>% 
    group_by(dpi, type, taxa_pair) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, taxa_pair, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "*", TRUE ~ ""))
w1_taxa %>% filter(p_label == "*")

## Kruskal-Wallis test for taxa pairs
kw_taxa <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ taxa_pair)

dunn_taxa <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    dunn_test(fractional_change ~ taxa_pair, p.adjust.method = "holm", detailed = TRUE) %>% 
    filter(p.adj < 0.05)

## Two samples Wilcoxon  test for taxa pairs over time
w2_dpi_taxa <- auc_fold_change %>% 
    group_by(type, taxa_pair) %>% 
    wilcox_test(fractional_change ~ dpi, p.adjust.method = "holm", detailed = TRUE)

```

```{r, include = FALSE}
## Kruskal Wallis test for strain pairs
kw_strain <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ strain_pair)

dunn_strain <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    dunn_test(fractional_change ~ strain_pair, p.adjust.method = "holm", detailed = TRUE) %>% 
    select(dpi, type, group1, group2, p, p.adj, p.adj.signif)

## Two samples Wilcoxon test for strain pairs
w2_strain <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ strain_pair, p.adjust.method = "holm", detailed = TRUE)
w2_strain %>% filter(p.adj < 0.05)

## One sample Wilcoxon test for strain pairs
w1_strain  <-  auc_fold_change %>% 
    group_by(dpi, strain_pair, type) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, p.adjust.method = "holm", detailed = TRUE) %>% 
    select(dpi, strain_pair, type, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "< 0.05", TRUE ~ as.character(p)))
w1_strain %>% filter(p < 0.05)
```

```{r, include = FALSE}
## Kruskal Wallis test for  Syncom
kw_syncom <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    kruskal_test(fractional_change ~ syncom)

## Two samples Wilcoxon test for SynCom
w2_syncom <- auc_fold_change %>% 
    group_by(type, dpi) %>% 
    wilcox_test(fractional_change ~ syncom, p.adjust.method = "holm", detailed = TRUE) %>% 
    filter(p.adj < 0.05)

## One sample Wilcoxon test for SynCom
w1_syncom <-  auc_fold_change %>% 
    group_by(dpi, type, syncom) %>% 
    wilcox_test(fractional_change ~ 1, mu = 0, detailed = TRUE) %>% 
    select(dpi, type, syncom, estimate, statistic, p) %>% 
    mutate(
        p_size = case_when(p < 0.05 ~ 0.05, TRUE ~ p),
        p_label = case_when(p < 0.05 ~ "*", TRUE ~ as.character(p)))
w1_syncom %>% filter(p_label == "*")
```

```{r, include = FALSE}
plt.frq <- fractions %>% 
    group_by(taxa_pair, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(x=r, y=mean_fraction))+
    facet_nested(taxa_pair ~ dpi + synID, 
                 nest_line = element_line(color="black"),
                 labeller = labeller(dpi = dpi.lab2))+
    geom_area(aes(fill=fct_reorder(type, index)), color = "black")+
    scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                       expand = c(0,0), limit=c(0.2,30), breaks = seq(5, 25, 10))+
    scale_y_continuous(name = expression(paste("Relative frequency of g(", italic(r),")")), 
                       expand = c(0,0), breaks = seq(0,1,0.5))+
    scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
    theme_rs()+
    theme(
          panel.spacing.x = grid::unit(0.5, "line"),
          strip.text.x = element_text(margin = margin(b = 1, t = 5), face = "plain"),
          strip.text.y = element_text(angle=0, vjust = 0.5, hjust = 0, face = "plain"),
          legend.position = "bottom")

plt.pattern <- left_join(auc_fold_change, w1_taxa, by = c("dpi", "type", "taxa_pair")) %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3)) %>% 
    ggplot(aes(taxa_pair, fractional_change))+
    geom_text(aes(y=2.5, label=p_label), size = 5, parse=FALSE)+
    facet_grid(dpi ~ fct_reorder(type, index), labeller = labeller(.rows = dpi.lab2, .cols = pattern.lab))+
    geom_jitter(color = "black", stroke = 0.5, width = 0.1, alpha = 0.8)+
    geom_boxplot(fill="white", alpha = 0.5, width = 0.2, outlier.alpha = 0)+
    geom_hline(yintercept = 0, linetype = "dashed")+
    theme_rs()+
    theme(strip.text = element_text(face = "plain"))+
    labs(x = "Taxa pair", y = "Fractional change")

```

```{r figure_main_6, fig.id=TRUE, dev = "png", fig.dim=c(5,8)}
plt.frq / wrap_elements(plot = plt.pattern) + 
    plot_annotation(tag_levels = "A") &
    theme(legend.box.just = "center",
          plot.margin = margin(0,1,0,1))
```


```{r plot_sup_freq_all, include=FALSE}
fractions_pairs <- fractions %>% 
    group_by(strain_pair, syncom, synID, dpi, type, r) %>% 
    summarise(mean_fraction = mean(fraction), .groups="drop") %>% 
    mutate(index = case_when(
        type == "regular_fraction" ~ 1,
        type == "random_fraction" ~ 2,
        type == "aggregate_fraction" ~ 3))

freq_plots <- function(name){
    plt <- fractions_pairs %>% 
        filter(strain_pair == name) %>% 
        ggplot(aes(x=r, y=mean_fraction))+
        facet_nested(synID + syncom ~ dpi, 
                     nest_line = element_line(color="black"),
                     labeller = labeller(dpi=dpi.lab2))+
        geom_area(aes(fill=fct_reorder(type, index)), color = "black")+
        scale_x_continuous(name = expression(paste("Distance, ", italic(r), " (", mu,"m)")), 
                           expand = c(0,0), limit=c(0.2,30), breaks = seq(10, 20, 10))+
        scale_y_continuous(name = expression(paste("Relative frequency of g(", italic(r),")")), 
                           expand = c(0,0), breaks = seq(0,1,0.5))+
        scale_fill_manual(name = "Spatial Pattern", labels = pattern.lab, values = pattern.pal)+
        theme_rs()+
        labs(title = pair.lab[name])+
        theme(panel.spacing.x = grid::unit(0.5, "line"),
              panel.spacing.y = grid::unit(1, "line"),
              strip.text = element_text(angle=0, vjust = 0.5, hjust = 0, 
                                          margin = margin(r = 5, l = 2), face = "plain"))
    return (plt)
}

list_plot <- unique_pair %>% 
    map(~freq_plots(.x))

```

```{r figure_supplemental_6, warning=FALSE, fig.id=TRUE, fig.cap="Frequency plots g", dev = "png", fig.dim=c(16,8)}
wrap_plots(list_plot, ncol = 5, guides = "collect")+
    plot_annotation(tag_levels = "A")
```
