---
title: "Bacterial community complexity in the phyllosphere penalises specialists over generalists - Results Section 1"
author: "Rudolf Schlechter"
output:
  html_document:
    df_print: paged
    keep_md: yes
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center', dpi = 300, cache = TRUE)
options(digits = 2, scipen = 1, pillar.signif = 3)

# Libraries
library(here)
library(tidyverse)
library(car)
library(rstatix)

# Plotting 
library(ggdist)
library(ggpubr)
library(patchwork)

# Other dependencies
source(here('code', 'palette_syncom.R'))
source(here('code', 'theme_rs_spatial.R'))
source(here('code', 'dunnet_function.R'))

set.seed(19900725)

```

## Taxon-specific population density changes correlate with community complexity

```{r import_data, include=FALSE}
# Load data
data_cfu <- read.csv(here("results", "data_processed.csv")) %>% 
    na.omit %>% 
    mutate(dpi = factor(dpi),
           synID = factor(synID),
           taxa = factor(taxa))
```

```{r cfu_str}
data_cfu %>% head
```

```{r}
# Linear regression for testing normality and homoscedasticity
linear_cfu = lm(cfu_log ~ synID + dpi + taxa, data_cfu)

# Shapiro-Wilk test for normality
cfu_normality = shapiro.test(rstandard(linear_cfu)); cfu_normality

# Breusch-Pagan test for homogeneity of variances
cfu_homoskedasticity = ncvTest(linear_cfu); cfu_homoskedasticity
```

We explored the impact of community complexity on bacterial population densities *in planta*, focusing on two prominent bacterial taxa, *Methylobacterium* and *Sphingomonas*. Employing a full factorial design, we evaluated bacterial population densities across varying levels of community complexities, characterised by three conditions: near-isogenic control (C), a two-species SynCom (S2), and a three-species SynCom (S3) at 7- and 14-days post-inoculation (dpi).

We used non-parametric methods to analyse the CFU data, considering violation of normality (Shapiro-Wilk test, *W* = `r cfu_normality$statistic`, *p* = `r cfu_normality$p.value`) and homogeneity of variance (Breusch-Pagan test, $\ X^{2}$ = `r cfu_homoskedasticity$ChiSquare`, *p* = `r cfu_homoskedasticity$p`)


```{r cfu_analysis_synID}
# Kruskal-Wallis test and effect size for community complexity (synID)
kw_synID = data_cfu %>% 
    kruskal_test(cfu_log ~ synID) %>% 
    mutate(p_label = case_when(
        p < 0.05 ~ "< 0.05", 
        TRUE ~ as.character(p)))

keff_synID = data_cfu %>% 
    kruskal_effsize(cfu_log ~ synID, nboot=100)

# Dunn's Test
dunn_synID = data_cfu %>% 
    dunn_test(cfu_log ~ synID, p.adjust.method = "holm") %>% 
    mutate(p_label = case_when(
        p.adj < 0.05 ~ "< 0.05", 
        TRUE ~ as.character(p.adj)))

# Fold change of population density by SynCom complexity (synID)
fc_cfu_synID = data_cfu %>% 
    # Group by SynCom (C, S2, S3)
    group_by(synID) %>% 
    summarise(median_cfu = median(cfu)) %>% 
    # Calculate the fold change (FC) in relation to C
    mutate(FC = median_cfu/median_cfu[1],
           logFC = log2(FC))
```

We first tested the influence of community complexity on individual bacterial populations at the CFU level, showing a pronounced and statistically significant effect (Kruskal-Wallis, *H*(`r kw_synID$df`) = `r kw_synID$statistic`, *p* `r kw_synID$p_label`). This effect was reflected by a significant `r fc_cfu_synID$FC[fc_cfu_synID$synID=="S2"]`-fold increase in population densities within two-species communities (Dunn's test, *Z* = `r abs(dunn_synID$statistic[1])`, *p*-adjusted `r dunn_synID$p_label[1]`), and a pronounced `r (fc_cfu_synID$FC[fc_cfu_synID$synID=="S3"])^-1`-fold reduction in the three-species communities (Dunn's test, *Z* = `r abs(dunn_synID$statistic[2])`, *p*-adjusted `r dunn_synID$p_label[2]`), relative to the near-isogenic control.


```{r cfu_analysis_dpi}
# Wilcoxon test and effect size for sampling time (dpi)
w_dpi = data_cfu %>% 
    wilcox_test(formula = cfu_log ~ dpi) %>% 
    mutate(p_label = case_when(
        p < 0.05 ~ "< 0.05", 
        TRUE ~ as.character(p)))

weff_dpi = data_cfu %>% 
    wilcox_effsize(formula = cfu_log ~ dpi, ci = TRUE, nboot = 100)

# Fold change of population density by time of sampling (dpi)
fc_cfu_dpi = data_cfu %>% 
    # Group by day (7, 14 dpi)
    group_by(dpi) %>% 
    summarise(median_cfu = median(cfu)) %>% 
    # Calculate the fold change (FC) in relation to 7 dpi
    mutate(FC = median_cfu/median_cfu[1],
           logFC = log2(FC))
```

Subsequently, we considered the temporal changes as an influencing factor in population density and evaluated how population density changed between the two sampling points. Here we observed a small yet significant temporal effect on population density (Wilcoxon, *W* = `r w_dpi$statistic`, *p* `r w_dpi$p_label`), with a `r fc_cfu_dpi$FC[fc_cfu_dpi$dpi=="14dpi"]`-fold increase between 7 and 14 dpi.

```{r cfu_analysis_taxa}
# Wilcoxon test and effect size for bacterial group (taxa)
w_taxa = data_cfu %>% 
    wilcox_test(formula = cfu_log ~ taxa) %>% 
    mutate(p_label = case_when(
        p < 0.05 ~ "< 0.05", 
        TRUE ~ as.character(p)))

weff_taxa = data_cfu %>% 
    wilcox_effsize(formula = cfu_log ~ taxa, ci = TRUE, nboot = 100)

# Fold change of population density by bacterial group (taxa)
fc_cfu_taxa = data_cfu %>% 
    # Group by taxa (Sphingomonas, Methylobacterium)
    group_by(taxa) %>% 
    summarise(median_cfu = median(cfu)) %>% 
    # Calculate the fold change (FC) in relation to Methylobacterium
    mutate(FC = median_cfu/median_cfu[1],
           logFC = log2(FC))
```

```{r, include=FALSE}
## Fold change and Dunn tests for two variables

## Variables
explanatory = "synID"
taxa_dpi = c("taxa", "dpi")
strain_dpi = c("strain", "dpi")

# Fold change of population density by taxa and dpi
dunntest_taxa_dpi <- dun_func1(data_cfu, "cfu_log", taxa_dpi, explanatory) %>% 
    mutate(symbol = ifelse(p_label == "< 0.05", "*", "")) %>% 
    rename(synID = group2)
fold_taxa_dpi <- fold_func1(data_cfu, "cfu", taxa_dpi, explanatory)

# Fold change of population density by strain and dpi
dunntest_strain_dpi <- dun_func1(data_cfu, "cfu_log", strain_dpi, explanatory) %>% 
    mutate(symbol = ifelse(p_label == "< 0.05", "*", "")) %>% 
    rename(synID = group2)
fold_strain_dpi <- fold_func1(data_cfu, "cfu", strain_dpi, explanatory)
```

Next, we evaluated the changes in populations of specific bacterial taxa, and found that *Methylobacterium* population densities were notably different to *Sphingomonas* (Wilcoxon, *Z* = `r w_taxa$statistic`, *p* `r w_taxa$p_label`). *Sphingomonas* population densities were `r fc_cfu_taxa$FC[fc_cfu_taxa$taxa=="Sphingomonas"]` times larger than those of *Methylobacterium*. Within *Sphingomonas*, SmFR1 consistently increased population sizes in S2, irrespective of the presence of a second species, and to a lesser extent in S3. By contrast, SpFA2 population sizes showed a transient increase in S2, but predominantly decreased in S3.

*Methylobacterium* populations responded negatively to increasing community complexity. MeL85, MeL92, and Mr0-1 consistently experienced a reduction in population sizes, particularly within S3 and, to a lesser extent, in S2. Among the *Methylobacterium* species, MeL92 and Mr0-1 benefited from the presence of any other species (S2). However, this effect was only observed at 14 dpi, and it was lost in the presence of a third competitor (S3). MeL85 was the most susceptible to population decrease over time, and across varying community complexities and compositions.

Collectively, these findings indicate that bacterial taxa differentially responded to community complexity within the leaf environment. Notably, *Methylobacterium* populations were more susceptible compared to sphingomonads. Among the studied species, *Sphingomonas* FR1 emerged as the most competitive, while *Methylobacterium* L85 was the least competitive.

## Figures

```{r figure_main_3_setup, include = FALSE}
# Plot CFU
# Multiple comparisons
plt.cfu.stat = dunn_synID %>% 
    mutate(p.adj.signif = case_when(
        p.adj.signif == "****" ~ "*", 
        TRUE ~ p.adj.signif), y.position = c(9,10,9.5))

# Plot A
plt.a <- data_cfu %>% 
    ggplot(aes(synID, cfu_log))+
    stat_eye(
        side="right",
        adjust = 1,
        justification = -0.3,
        .width = 0,
        scale = 0.5,
        point_colour = NA)+
    geom_jitter(width = 0.1, alpha = 0.1)+
    geom_boxplot(fill = "white", width=0.2, outlier.alpha = 0)+
    coord_cartesian(xlim=c(1, 3.2))+
    stat_pvalue_manual(
        data = plt.cfu.stat,
        size = 4,
        label = "p.adj.signif",
        xmin = "group1", xmax = "group2",
        y.position = "y.position")+
    scale_y_continuous(limits = c(3.5, 10.2), breaks = c(4, 6, 8, 10))+
    labs(y = plt_bac_density_lab,
         x = "SynCom")+
    theme_rs()+
    theme(axis.text.x = element_text(hjust = 0.5, vjust = 1.5))

# Plot B
plt.b <- inner_join(dunntest_taxa_dpi, fold_taxa_dpi, by = c("taxa", "dpi", "synID")) %>% 
    filter(group1 == "C") %>% 
    ggplot(aes(x = synID, y = taxa))+
    facet_wrap(~dpi, ncol = 2, labeller = labeller(dpi = dpi.lab2))+
    geom_tile(aes(fill = log2FC), colour = "black")+
    geom_text(aes(label = symbol), vjust = 0.77, size = 5)+
    #coord_equal(ratio = 0.5)+
    scale_fill_gradientn(name = bquote(Log[2]~"FC"), 
                         colours = wes_palette("Zissou1")[c(1, 2, 3, 5)], 
                         values = c(0, 0.55, 1), 
                         limits = c(-8, 4), 
                         breaks = seq(-8, 4, 4), 
                         na.value = 'grey90')+
    scale_y_discrete(name = "", labels = taxa.lab)+
    labs(x = "")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust = 0.5, vjust = 3),
          axis.text.y = element_text(face = "italic"),
          strip.text = element_text(face = "plain"),
          legend.position = "bottom")

# Plot C
plt.c <- inner_join(dunntest_strain_dpi, 
                     fold_strain_dpi, 
                     by = c("strain", "dpi", "synID")) %>% 
    filter(group1 == "C") %>% 
    ggplot(aes(x = synID, y = strain))+
    facet_wrap(~dpi, ncol = 2, labeller = labeller(dpi=dpi.lab2))+
    geom_tile(aes(fill = log2FC), colour = "black")+
    geom_text(aes(label = symbol), vjust = 0.77, size = 5)+
    #coord_equal(ratio = 0.5)+
    scale_fill_gradientn(name = bquote(Log[2]~"FC"), 
                         colours = wes_palette("Zissou1")[c(1,2,3,5)], 
                         values = c(0,0.55,1), 
                         limits = c(-8,4), 
                         breaks = seq(-8,4,4), 
                         na.value = 'grey90')+
    scale_y_discrete(name = "Focal strain", labels = sp.lab)+
    scale_x_discrete(name = "SynCom")+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(hjust = 0.5, vjust = 3),
          strip.text = element_text(face = "plain"),
          legend.position = "bottom")
```

```{r figure_2, fig.id=TRUE, fig.cap="Bacterial population density in the arabidopsis phyllosphere", fig.width=3.5}

wrap_plots(wrap_elements(plot = plt.a), plt.b, plt.c, ncol = 1) + 
    plot_annotation(tag_levels = "A") + 
    plot_layout(guides = "collect",
                heights = c(6,1,2)) & 
    theme(legend.box.just = "left",
          legend.position = "bottom",
          panel.spacing.x = unit(0.5, "lines"),
          plot.margin = margin(1,1,0,1),
          plot.tag = element_text(size = 7)) &
    guides(fill = guide_colorbar(title.position="top", title.hjust = 0.5, barheight = unit(0.1, 'in')))

# Save plot
ggsave(here("results", "figure_2.png"), width = 3.5, height = 5)
ggsave(here("results", "figure_2.tiff"), width = 3.5, height = 5)

```

# Supplemental Figure

```{r figure_sup_1_setup, include=FALSE}
dunntest_syncom <- dun_func1(data_cfu, "cfu_log", c("strain", "dpi"), "syncom") %>% 
    separate(group1, into=c('synID', 'com')) %>% 
    mutate(symbol = ifelse(p_label == "< 0.05", "*", "")) %>% 
    rename(syncom = group2) %>% 
    group_by(strain, dpi) %>% 
    filter(synID=="C") %>% 
    select(-synID:-com)

dunn_S2 = dunntest_syncom %>% 
    separate(syncom, into = c('synID', 'com'), remove = FALSE) %>% 
    filter(synID == "S2") %>% 
    select(syncom, dpi, strain, symbol) %>% 
    group_by(syncom, dpi) %>% 
    mutate(to = rev(strain)) %>% 
    ungroup() %>% 
    select(syncom, dpi, strain, to, symbol)

dunn_test_long <- dunntest_syncom %>% 
    separate(syncom, into = c('synID', 'com'), remove = FALSE) %>% 
    filter(synID == "S3") %>% 
    mutate(to = syncom) %>% 
    select(syncom, dpi, strain, to, symbol) %>% 
    rbind(dunn_S2,.) %>% 
    select(dpi:symbol) %>% 
    rename(name=to)

data_median_control = data_cfu %>% 
    filter(synID == "C") %>% 
    group_by(strain, dpi) %>% 
    summarise(m_mono = median(cfu), .groups="drop")

data_fold = data_cfu %>% 
    group_by(synID, syncom, strain, dpi) %>% 
    summarise(m = median(cfu), .groups="drop") %>% 
    inner_join(., data_median_control, by = c("strain", "dpi")) %>%
    filter(synID != "C") %>% 
    mutate(fold = log2(m/m_mono)) %>% 
    left_join(., dunntest_syncom, by = c("strain", "syncom", "dpi"))

data_fold_S2 = data_fold %>% 
    filter(synID == "S2") %>% 
    select(syncom, strain, fold, dpi) %>% 
    group_by(syncom, dpi) %>% 
    mutate(to=rev(strain)) %>% 
    pivot_wider(id_cols=c(strain, dpi), names_from = to, values_from = fold) %>% 
    select(strain, dpi, meL85, meL92, mr01, smfr1, spfa2) %>% 
    arrange(strain)

data_fold_S3 = data_fold %>% 
    filter(synID == "S3") %>% 
    pivot_wider(id_cols=c(strain, dpi), names_from = syncom, values_from=fold) %>% 
    arrange(strain)

```

```{r figure_s2, warning=FALSE, fig.dim=c(3.5,3)}

# Plot
plt.sup = cbind(data_fold_S2, data_fold_S3[,c(-1:-2)]) %>% 
    pivot_longer(cols = -c(strain,dpi)) %>% 
    ungroup %>% 
    mutate(label = if_else(grepl("S3.*", name), "S3", "S2"),
           index = seq(1:nrow(.))) %>% 
    left_join(., dunn_test_long, by = c("strain", "dpi", "name")) %>% 
    ggplot(aes(y = strain, x = fct_reorder(name, index)))+
    facet_grid(dpi~label, labeller = labeller(dpi = dpi.lab2), scales = "free_x", space = "free")+
    geom_tile(aes(fill = value), colour= "black")+
    geom_text(aes(label=symbol), vjust = 0.77, size = 5)+
    scale_fill_gradientn(name = bquote(Log[2]~"FC"), 
                         colours = wes_palette("Zissou1")[c(1,2,3,5,5)], 
                         values = c(0,0.57,1), limits=c(-10,8), 
                         breaks = seq(-8,4,4), na.value = 'grey90')+
    scale_y_discrete(name="Focal strain", labels = sp.lab)+
    scale_x_discrete(name="Competitor(s)", labels = sp.lab)+
    theme_rs()+
    theme(panel.border = element_blank(),
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          strip.text = element_text(face = "plain"),
          panel.spacing = unit(0.5, "lines"),
          legend.position = "bottom")+
    guides(fill = guide_colorbar(title.position="left", title.vjust = 0.75))

plt.sup

# Save plot  
ggsave(here("results", "figure_s2.png"), width = 3.5, height = 3)
ggsave(here("results", "figure_s2.tiff"), width = 3.5, height = 3)
```

